Racial and Ethnic Differences in ETI

### 2024-03-18 Jordana Hoppe/SSwAB-IT/phase_1_analysis.qmd
minor chnges

```{r setup}
#| include: false
library(tidyverse)
library(gtsummary)
library(redcapAPI)
library(irr)
library(gt)
library(caret)
```

```{r data cleaning}
#| include: false
# Import from REDCap
unlockREDCap(c(rcon = "SSwAB-IT (New)"),
  keyring = "API_KEYs",
  envir = 1,
  url = "https://redcap.ucdenver.edu/api/"
)
exportBulkRecords(list(db = rcon), envir = 1)
# Exclude
exclude <- c("01-030_W", "01-038_W", "02-0038", "02-0039", "04-001")
# Combine collection results
df <- full_join(db_cos_collection_results, db_sos_collection_results)
# Format
df <- df %>%
  select(
    record_id, redcap_data_access_group,
    staphylococcus_aureus_cos:other_result_cos,
    staphylococcus_aureus_sos:other_result_sos
  ) %>%
  filter(!record_id %in% exclude)
# Remove empty demographics rows
db_demographics <- db_demographics %>%
  filter(
    redcap_event_name == "Clinic Visit (Arm 1: Eligible)",
    !record_id %in% exclude
  )
# Excluded participants
db_medications <- db_medications %>% filter(!record_id %in% exclude)
# Combine race and ethnicity columns
races <- c(
  "American Indian or Alaskan Native", "Asian", "Hawaiian or Pacific Islander",
  "Black or African American", "White", "Unknown", "Other"
)
ethnicities <- c("Hispanic", "Non-Hispanic", "Unknown/Not Reported")
db_demographics$Race <- apply(db_demographics, 1, function(r) {
  w <- which(r[grep("race", names(r))] == "Checked")
  if (length(w) == 1) {
    return(races[w])
  } else if (length(w) > 1) {
    return("Multiple Races")
  } else if (length(w) == 0) {
    return(NA)
  }
})
db_demographics$Ethnicity <- apply(db_demographics, 1, function(r) {
  w <- which(r[grep("ethnicity", names(r))] == "Checked")
  if (length(w) == 1) {
    return(ethnicities[w])
  } else if (length(w) > 1) {
    return(paste(ethnicities[w], sep = "/"))
  } else if (length(w) == 0) {
    return(NA)
  }
})
# # Drop levels
# level_drop <- c("sex", "Race", "Ethnicity")
# db_demographics[, level_drop] <- lapply(
#   db_demographics[, level_drop],
#   function(c) {
#     droplevels(factor(c))
#   }
# )
# Add groups to other dataframes
db_medications <- left_join(db_medications,
  db_demographics %>% select(record_id, group),
  by = join_by(record_id)
)
# Bug list
bugs <- c(
  "staphylococcus_aureus", "haemophilus_influenza", "pseudomonas_aeruginosa",
  "pseudomonas_non_aero", "s_maltophilia", "burkholderia", "aspergillus", "ntm",
  "other"
)
bug_names <- c(
  "S. aureus", "H. influenza", "P. aeruginosa", "Pseudomonas non-aeruginosa",
  "S. maltophilia", "Burkholderia", "Aspergillus", "NTM", "Other"
)
# Convert yes/no to positive/negative
df[, paste0(bugs, "_cos")] <- lapply(df[, paste0(bugs, "_cos")], factor,
  levels = c("No", "Yes"), labels = c("Negative", "Positive")
)
df[, paste0(bugs, "_sos")] <- lapply(df[, paste0(bugs, "_sos")], factor,
  levels = c("No", "Yes"), labels = c("Negative", "Positive")
)
# Create overall columns (any bug detected in sample)
df$overall_cos <- apply(df, 1, function(r) {
  any(r[paste0(bugs, "_cos")] == "Positive")
})
df$overall_cos <- factor(df$overall_cos,
  levels = c(F, T), labels = c("Negative", "Positive")
)
df$overall_sos <- apply(df, 1, function(r) {
  any(r[paste0(bugs, "_sos")] == "Positive")
})
df$overall_sos <- factor(df$overall_sos,
  levels = c(F, T), labels = c("Negative", "Positive")
)
# Add overall to list
bugs <- c(bugs, "overall")
bug_names <- c(bug_names, "Overall")
bug_list <- as.list(bug_names)
names(bug_list) <- bugs
```

@@

Cohen's kappa is interpreted analagously to a correlation coefficient, with -1 indicating perfect disagreement and 1 indicating perfect agreement. Tables with NaN and NAs indicate that all tests had the same values (i.e. all absent or all present).

| Cohen's Kappa | Interpretation         |
|---------------|------------------------|
| 0             | No agreement           |
| 0.10 - 0.20   | Slight agreement       |
| 0.21 - 0.40   | Fair agreement         |
| 0.41 - 0.60   | Moderate agreement     |
| 0.61 - 0.80   | Substantial agreement  |
| 0.81 - 0.99   | Near perfect agreement |
| 1             | Perfect agreement      |

Given a 2x2 table of the form

| SOS      | COS      | n |
|----------|----------|---|
| Negative | Negative | D |
| Positive | Negative | B |
| Negative | Positive | C |
| Positive | Positive | A |

the accuracy metrics are calculated as follows:

$Sensitivity=\frac{A}{A+C}$

$Specificity=\frac{D}{B+D}$

$Prevalence=\frac{A+C}{A+B+C+D}$

$PPV=\frac{sensitivity∗prevalence}{(sensitivity∗prevalence)+((1−specificity)∗(1−prevalence))}$

$NPV=\frac{specificity∗(1−prevalence)}{((1−sensitivity)∗prevalence)+((specificity)∗(1−prevalence))}$

$DetectionRate=\frac{A}{A+B+C+D}$

$DetectionPrevalence=\frac{A+B}{A+B+C+D}$

$BalancedAccuracy=\frac{sensitivity+specificity}{2}$

$Precision=\frac{A}{A+B}$

$Recall=\frac{A}{A+C}$

$F1=\frac{2∗precision∗recall}{precision+recall}$

>>>>

```{r setup}
#| include: false
library(redcapAPI)
library(tidyverse)
library(gt)
library(gtsummary)
library(caret)
```

```{r data cleaning}
#| include: false
# Import from REDCap
unlockREDCap(c(rcon = "SSwAB-IT (New)"),
  keyring = "API_KEYs",
  envir = 1,
  url = "https://redcap.ucdenver.edu/api/"
)
exportBulkRecords(list(db = rcon), envir = 1)
# Exclude
exclude <- c("01-030_W", "01-038_W", "02-0038", "02-0039", "04-001")
# Combine collection results
df <- full_join(db_cos_collection_results, db_sos_collection_results)
# Format
df <- df %>%
  select(
    record_id, redcap_data_access_group,
    staphylococcus_aureus_cos:other_result_cos,
    staphylococcus_aureus_sos:other_result_sos
  ) %>%
  filter(!record_id %in% exclude)
# Remove empty demographics rows
db_demographics <- db_demographics %>%
  filter(
    redcap_event_name == "Clinic Visit (Arm 1: Eligible)",
    !record_id %in% exclude
  )
# Excluded participants
db_medications <- db_medications %>% filter(!record_id %in% exclude)
# Combine race and ethnicity columns
races <- c(
  "American Indian or Alaskan Native", "Asian", "Hawaiian or Pacific Islander",
  "Black or African American", "White", "Unknown", "Other"
)
ethnicities <- c("Hispanic", "Non-Hispanic", "Unknown/Not Reported")
db_demographics$Race <- apply(db_demographics, 1, function(r) {
  w <- which(r[grep("race", names(r))] == "Checked")
  if (length(w) == 1) {
    return(races[w])
  } else if (length(w) > 1) {
    return("Multiple Races")
  } else if (length(w) == 0) {
    return(NA)
  }
})
db_demographics$Ethnicity <- apply(db_demographics, 1, function(r) {
  w <- which(r[grep("ethnicity", names(r))] == "Checked")
  if (length(w) == 1) {
    return(ethnicities[w])
  } else if (length(w) > 1) {
    return(paste(ethnicities[w], sep = "/"))
  } else if (length(w) == 0) {
    return(NA)
  }
})
# # Drop levels
# level_drop <- c("sex", "Race", "Ethnicity")
# db_demographics[, level_drop] <- lapply(
#   db_demographics[, level_drop],
#   function(c) {
#     droplevels(factor(c))
#   }
# )
# Add groups to other dataframes
db_medications <- left_join(db_medications,
  db_demographics %>% select(record_id, group),
  by = join_by(record_id)
)
# Bug list
bugs <- c(
  "staphylococcus_aureus", "haemophilus_influenza", "pseudomonas_aeruginosa",
  "pseudomonas_non_aero", "s_maltophilia", "burkholderia", "aspergillus", "ntm",
  "other"
)
bug_names <- c(
  "S. aureus", "H. influenza", "P. aeruginosa", "Pseudomonas non-aeruginosa",
  "S. maltophilia", "Burkholderia", "Aspergillus", "NTM", "Other"
)
# Convert yes/no to positive/negative
df[, paste0(bugs, "_cos")] <- lapply(df[, paste0(bugs, "_cos")], factor,
  levels = c("No", "Yes"), labels = c("Negative", "Positive")
)
df[, paste0(bugs, "_sos")] <- lapply(df[, paste0(bugs, "_sos")], factor,
  levels = c("No", "Yes"), labels = c("Negative", "Positive")
)
# Create overall columns (any bug detected in sample)
df$overall_cos <- apply(df, 1, function(r) {
  any(r[paste0(bugs, "_cos")] == "Positive")
})
df$overall_cos <- factor(df$overall_cos,
  levels = c(F, T), labels = c("Negative", "Positive")
)
df$overall_sos <- apply(df, 1, function(r) {
  any(r[paste0(bugs, "_sos")] == "Positive")
})
df$overall_sos <- factor(df$overall_sos,
  levels = c(F, T), labels = c("Negative", "Positive")
)
# Add overall to list
bugs <- c("overall", bugs)
bug_names <- c("Across all bugs", bug_names)
bug_list <- as.list(bug_names)
names(bug_list) <- bugs
# Agreement columns
df[, bugs] <- lapply(bugs, function(b) {
  cos <- df[, paste0(b, "_cos")]
  sos <- df[, paste0(b, "_sos")]
  f <- factor(paste(cos, sos, sep = "_"),
    exclude = c(
      "NA_Negative", "Negative_NA", "NA_Positive", "Positive_NA", "NA_NA"
    )
  )
  f <- factor(f,
    levels = c(
      "Negative_Negative", "Negative_Positive", "Positive_Negative",
      "Positive_Positive"
    ),
    labels = c(
      "COS Negative / SOS Negative", "COS Negative / SOS Positive",
      "COS Positive / SOS Negative", "COS Positive / SOS Positive"
    )
  )
  return(f)
})
```

@@

```{r}
df %>%
  select(all_of(bugs)) %>%
  tbl_summary(label = bug_list, missing_text = "Missing") %>%
  add_ci(statistic=list(all_categorical() ~ "{conf.low} - {conf.high}%"))
```

### 2024-03-18 Kamyron Jordan/Racial and Ethnic Differences in ETI/racial_and_ethnic_differences_in_eti.qmd
minor chnges

```{r results='asis'}
invisible(lapply(c("2019", "2020", "2021", "2022"), function(n) {
  cat("\n")
  cat("###", n)
  cat("\n")
  cat("\n")
  cat("#### All patients")
  cat("\n")
  d <- years[[n]]
  t <- d %>%
    select(eti_elig, contains("Vx"), race) %>%
    tbl_summary(
      by = race, missing_text = "Missing",
      type = all_dichotomous() ~ "categorical"
    ) %>%
    add_p(test.args = all_tests("fisher.test") ~ list(simulate.p.value = T)) %>%
    separate_p_footnotes() %>%
    bold_labels()
  print(t)
  cat("\n")
  cat("\n")
  cat("#### Eligible and prescribed off-label only")
  cat("\n")
  t <- d %>%
    filter(eti_elig == "Yes" | Vx445comb == "Yes") %>%
    select(eti_elig, contains("Vx"), race) %>%
    tbl_summary(
      by = race, missing_text = "Missing",
      type = all_dichotomous() ~ "categorical"
    ) %>%
    add_ci() %>%
    add_p(test.args = all_tests("fisher.test") ~ list(simulate.p.value = T)) %>%
    separate_p_footnotes() %>%
    bold_labels()
  print(t)
  cat("\n")
}))
```

@@

```{r}
# Create survival dataset
surv_dat <- encounter %>%
  mutate(stop = as.numeric(encounterdate - eti_elig_date)) %>%
  filter(stop > 0) %>%
  group_by(eDWID) %>%
  mutate(start = lag(stop, default = 0)) %>%
  select(
    eDWID, race, Insurance, family_income, patient_education, father_education,
    mother_education, encounterdate, encounterage, eti_elig_date, Vx445comb,
    start, stop
  ) %>%
  filter(stop <= first(stop[Vx445comb == "Yes"]))
surv_dat$Vx445comb <- as.numeric(factor(surv_dat$Vx445comb, levels = c("No", "Yes"))) - 1
# Fit
y <- "Surv(time = start, time2 = stop, event = Vx445comb)"
surv_f <- as.formula(paste0(y, "~race"))
fit <- survfit(surv_f, id = eDWID, data = surv_dat)
```

@@

```{r}
# Plot
names(fit$strata) <- gsub("race=", "", names(fit$strata))
fit %>%
  ggsurvfit(type = "risk") +
  add_confidence_interval() +
  add_risktable(
    risktable_stats =
      c("{round(n.risk)}", "{round(cum.event)}")
  ) +
  add_risktable_strata_symbol() +
  add_quantile() +
  ylab("Proportion on Vx445comb") + xlab("Days From Eligibility")
```

@@

```{r}
fit <- survfit(surv_f, id = eDWID, data = surv_dat[surv_dat$race != "Asian", ])
# Plot
names(fit$strata) <- gsub("race=", "", names(fit$strata))
fit %>%
  ggsurvfit(type = "risk") +
  add_confidence_interval() +
  add_risktable(
    risktable_stats =
      c("{round(n.risk)}", "{round(cum.event)}"),
    stats_label = c("Eligible", "Num. Started")
  ) +
  add_risktable_strata_symbol() +
  add_quantile() +
  ylab("Proportion on Vx445comb") + xlab("Days From Eligibility")
```

@@

```{r}
cox_res <- coxph(surv_f, data = surv_dat, cluster = eDWID)
cox_res %>%
  tbl_regression() %>%
  add_n(location = "level") %>%
  add_nevent(location = "level")
ggcoef_model(cox_res,
  show_p_values = F, add_reference_rows = F,
  signif_stars = F, colour = F,
)
```

#### Adjusted for insurance status, family income, and education

```{r}
new_surv_f <- update(surv_f, . ~ . + Insurance + family_income +
  patient_education + father_education + mother_education)
cox_res <- coxph(new_surv_f, data = surv_dat, cluster = eDWID)
cox_res %>%
  tbl_regression() %>%
  add_n(location = "level") %>%
  add_nevent(location = "level")
ggcoef_model(cox_res,
  show_p_values = F, add_reference_rows = F,
  signif_stars = F, colour = F,
)
```

@@

#### Adjusted for insurance status only

```{r}
new_surv_f <- update(surv_f, . ~ . + Insurance)
cox_res <- coxph(new_surv_f, data = surv_dat, cluster = eDWID)
cox_res %>%
  tbl_regression() %>%
  add_n(location = "level") %>%
  add_nevent(location = "level")
ggcoef_model(cox_res,
  show_p_values = F, add_reference_rows = F,
  signif_stars = F, colour = F,
)
```

@@

State special needs program, e.g., BCMH, CCS, CRS, GHPP, etc. = Other

@@

```{r}
surv_dat$Insurance <- factor(surv_dat$Insurance,
  levels = c(
    "Health Insurance Policy (e.g. Private Insurance)",
    "Indian Health Service",
    "Medicaid",
    "Medicare",
    "More than one",
    "No Insurance",
    "Other",
    "State special needs program, e.g., BCMH, CCS, CRS, GHPP, etc.",
    "TriCare or other military health plan"
  ),
  labels = c(
    "Private", "Public", "Public", "Public", "More Than One", "No Insurance",
    "Other", "Other", "Tricare"
  )
)
cox_res <- coxph(new_surv_f, data = surv_dat, cluster = eDWID)
cox_res %>%
  tbl_regression() %>%
  add_n(location = "level") %>%
  add_nevent(location = "level")
ggcoef_model(cox_res,
  show_p_values = F, add_reference_rows = F,
  signif_stars = F, colour = F,
)
```

# People without visits in a given year

```{r}
# Which are missing from encounters data?
a <- as.data.frame.matrix(table(annual$eDWID, annual$ReviewYear))
a <- a %>% rownames_to_column("eDWID")
e <- as.data.frame.matrix(table(by_year$eDWID, by_year$reviewyear))
e <- e %>% rownames_to_column("eDWID")
rowdiff_a <- setdiff(a, e)
rowdiff_a$data <- "annual"
rowdiff_e <- setdiff(e, a)
rowdiff_e$data <- "encounter"
diffs <- rbind(rowdiff_a, rowdiff_e)
diffs <- diffs %>%
  arrange(eDWID, desc(data)) %>%
  select(-data)
diffs <- diffs %>%
  group_by(eDWID) %>%
  reframe(across(`2015`:`2021`, ~ diff(as.numeric(.x))))
diffs %>%
  reframe(across(`2015`:`2021`, ~ sum(.x == 1))) %>%
  gt()
```

There were a total of `r length(setdiff(a$eDWID,e$eDWID))` included in the annual dataset that did not appear in the encounter-level dataset at all. There were an additional `r nrow(diffs)` patients who were included in both datasets, but had at least one year with no encounters. The table above shows the number of people missing an encounter in each year.

If someone was not seen a given year, do we count them as eligible for that year? I am assuming they will be eligible but obviously will not have started any modulators.

# Potential COVID effects

## Spaghetti plot

```{r}
#| warning: false
ggplot(encounter, aes(
  y = GLI_FEV1_pct_predicted, x = encounterdate,
  group = eDWID, color = factor(eDWID)
)) +
  geom_line(alpha = 0.02) +
  theme_bw() +
  theme(legend.position = "none")
```

## Visits prior to and after March 11, 2020

The table below is limited to visits 2019 - 2021.

```{r}
encounter %>%
  filter(encounterdate >= "2019-01-01") %>%
  mutate(post_covid = encounterdate > "2020-03-11") %>%
  group_by(eDWID) %>%
  summarise(
    `Number of Visits Pre-COVID` = sum(!post_covid),
    `Number of Visits Post-COVID` = sum(post_covid),
    race = first(race),
    .groups = "drop"
  ) %>%
  select(-eDWID) %>%
  tbl_summary(by = race, missing_text = "Missing") %>%
  bold_labels()
```

# Lung function

## Annualized averages

FEV1 and FVC percent predicted (based on GLI equations) were averaged by year for each person.

### FEV1

#### Basic model

The most basic model we can fit to these data uses the average values as the outcome, with terms for review year, race, and their interaction and a random intercept for each patient. This model is adjusted for age, private vs. other insurance, and parental education.

```{r}
# Most basic model of lung function over time by race
# Try a few different models
mod <- lmer(
  mean_ppFEV ~ factor(ReviewYear) * race + (1 | eDWID),
  data = annual
)
res <- tidy(mod, effects = "fixed")
means <- data.frame(emmeans(mod, ~ race | ReviewYear,
  lmerTest.limit = 1e6,
  pbkrtest.limit = 1e6
))
```

>>>>

```{r results='asis'}
invisible(lapply(c("2020", "2021", "2022"), function(n) {
  cat("\n")
  cat("###", n)
  cat("\n")
  cat("\n")
  cat("#### All patients")
  cat("\n")
  d <- years[[n]]
  t <- d %>%
    select(eti_elig, contains("Vx"), race) %>%
    tbl_summary(
      by = race, missing_text = "Missing",
      type = all_dichotomous() ~ "categorical"
    ) %>%
    add_p(test.args = all_tests("fisher.test") ~ list(simulate.p.value = T)) %>%
    separate_p_footnotes() %>%
    bold_labels()
  print(t)
  cat("\n")
  cat("\n")
  cat("#### Eligible and prescribed off-label only")
  cat("\n")
  t <- d %>%
    filter(eti_elig == "Yes" | Vx445comb == "Yes") %>%
    select(eti_elig, contains("Vx"), race) %>%
    tbl_summary(
      by = race, missing_text = "Missing",
      type = all_dichotomous() ~ "categorical"
    ) %>%
    add_ci() %>%
    add_p(test.args = all_tests("fisher.test") ~ list(simulate.p.value = T)) %>%
    separate_p_footnotes() %>%
    bold_labels()
  print(t)
  cat("\n")
}))
```

@@

```{r}
# Create survival dataset
surv_dat <- encounter %>%
  mutate(stop = as.numeric(encounterdate - eti_elig_date)) %>%
  filter(stop > 0) %>%
  group_by(eDWID) %>%
  mutate(start = lag(stop, default = 0)) %>%
  select(
    eDWID, race, Insurance, family_income, patient_education, father_education,
    mother_education, encounterdate, encounterage, gli_fev1_ppred_rn,
    eti_elig_date, Vx445comb, start, stop
  ) %>%
  filter(stop <= first(stop[Vx445comb == "Yes"]))
surv_dat$Vx445comb <- as.numeric(factor(surv_dat$Vx445comb, levels = c("No", "Yes"))) - 1
# Fit
y <- "Surv(time = start, time2 = stop, event = Vx445comb)"
surv_f <- as.formula(paste0(y, "~race"))
fit <- survfit(surv_f, id = eDWID, data = surv_dat)
```

@@

```{r}
#| warning: false
# Plot
names(fit$strata) <- gsub("race=", "", names(fit$strata))
fit %>%
  ggsurvfit(type = "risk") +
  add_confidence_interval() +
  add_risktable(
    risktable_stats =
      c("{round(n.risk)}", "{round(cum.event)}")
  ) +
  add_risktable_strata_symbol() +
  add_quantile() +
  ylab("Proportion on ETI") + xlab("Days From Eligibility")
```

@@

```{r}
#| warning: false
fit <- survfit(surv_f, id = eDWID, data = surv_dat[surv_dat$race != "Asian", ])
# Plot
names(fit$strata) <- gsub("race=", "", names(fit$strata))
fit %>%
  ggsurvfit(type = "risk") +
  add_confidence_interval() +
  add_risktable(
    risktable_stats =
      c("{round(n.risk)}", "{round(cum.event)}"),
    stats_label = c("Eligible", "Num. Started")
  ) +
  add_risktable_strata_symbol() +
  add_quantile() +
  ylab("Proportion on ETI") + xlab("Days From Eligibility")
```

@@

```{r}
cox_res <- coxph(surv_f, data = surv_dat, cluster = eDWID)
cox_res %>%
  tbl_regression() %>%
  add_n(location = "level") %>%
  add_nevent(location = "level")
ggcoef_model(cox_res,
  show_p_values = F, add_reference_rows = F,
  signif_stars = F, colour = F, exponentiate = T
)
```

@@

@@

State special needs program, e.g., BCMH, CCS, CRS, GHPP, etc. = Public

@@

```{r}
surv_dat$Insurance <- factor(surv_dat$Insurance,
  levels = c(
    "Health Insurance Policy (e.g. Private Insurance)",
    "Indian Health Service",
    "Medicaid",
    "Medicare",
    "More than one",
    "No Insurance",
    "Other",
    "State special needs program, e.g., BCMH, CCS, CRS, GHPP, etc.",
    "TriCare or other military health plan"
  ),
  labels = c(
    "Private", "Public", "Public", "Public", "More Than One", "No Insurance",
    "Other", "Public", "Tricare"
  )
)
new_surv_f <- update(surv_f, . ~ . + Insurance)
cox_res <- coxph(new_surv_f, data = surv_dat, cluster = eDWID)
cox_res %>%
  tbl_regression() %>%
  add_n(location = "level") %>%
  add_nevent(location = "level")
ggcoef_model(cox_res,
  show_p_values = F, add_reference_rows = F,
  signif_stars = F, colour = F, exponentiate = T
)
```

#### Adjusted for insurance, age group, and FEV1

```{r}
surv_dat$`Age Group` <- cut(surv_dat$encounterage,
  breaks = c(-Inf, 6, 12, 18, 30, Inf), right = F
)
new_surv_f <- update(surv_f, . ~ . + Insurance + `Age Group` +
  gli_fev1_ppred_rn)
cox_res <- coxph(new_surv_f, data = surv_dat, cluster = eDWID)
cox_res %>%
  tbl_regression() %>%
  add_n(location = "level") %>%
  add_nevent(location = "level")
ggcoef_model(cox_res,
  show_p_values = F, add_reference_rows = F,
  signif_stars = F, colour = F, exponentiate = T
)
```
